On the Synchronization of Computational RFIDs

### 2018-06-29 TMC-2017-11-0695.R1-main.tex
minor revision

\author{Kas{\i}m~Sinan~Y{\i}ld{\i}r{\i}m, 
	Henko~Aantjes, 
	Przemys{\l}a~Pawe{\l}czak~\IEEEmembership{Member,~IEEE,} 
	and~Amjad~Yousef~Majid
	%
	\IEEEcompsocitemizethanks{
		\IEEEcompsocthanksitem A preliminary version of this article was presented at the HLPC 2016 Workshop, Atlanta, GA, USA, April 2, 2016~\cite{hlpc2016}. \protect\\
		\IEEEcompsocthanksitem The authors are with the Embedded Software Group, Delft University of Technology, Mekelweg 4, 2628 CD Delft, The Netherlands (e-mail: \{k.s.yildirim, p.pawelczak, a.y.majid\}@tudelft.nl, h.aantjes@student.tudelft.nl).
	}
%	\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}
}

@@

As of now, battery-less networks are in the form of one-hop architecture in which the CRFID tags communicate only with the RFID reader~\cite{wisp,booksmith2013,wisent:2016}. \add{Therefore, CRFID tags cannot operate in a multi-hop manner.} However, recent advancements in the wireless communication enabled tag-to-tag communication, e.g.~\cite{liu2013ambient}.  We anticipate that \add{this communication capability will lead up multi-hop CRFID networks that} will demand their own implementation of basic sensor network building blocks~\cite{levis:tinyos} with the aim of replacing the existing battery-powered sensing and data collection applications with their battery-less counterparts.

\add{Since we are not aware of any network infrastructure demonstrating neither multi-hop communication nor multi-hop synchronization, our focus in this article are the CRFID networks where several readers cover a tag population simultaneously in a single hop. A representative application is the greenhouse monitoring where several tags are powered by multiple readers, see Figure~\ref{fig:motivation}. In this application, the tags sense the environment and timestamp their temperature readings. When the readers request their temperature readings, the tags transmit their timestamps together with the collected temperature data. In order to make inferences on the collected temperature readings, timestamps should be consistent so that the readers can provide a temporal ordering of the readings. Therefore, tags require a common time notion to timestamp their readers---creating a demand for explicit time synchronization. In order to achieve synchronization, readers should first acquire the global time, e.g. by accessing time servers using the Internet; then individual tags should be synchronized to the individual readers.}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.75\columnwidth]{figures/scenario.pdf}
	\caption{\label{fig:motivation}\color{blue}A representative greenhouse monitoring application using CRFIDs. Two readers are deployed to provide power to the individual tags and to collect their periodic sensor readings.}
\end{figure}

@@

\add{The EPC Gen2 protocol uses slotted Aloha that performs a form of synchronization between the reader and the tags: by adjusting clock offset differences and introducing guard times, tags are ensured to communicate at the allocated slot boundaries. However, this method does not compensate for clock drifts and output a synchronized clock value that can be accessed by the tags to timestamp (and make inference about) the collected information, e.g. from the sensors attached to tags. In consequence, applications requiring time notion cannot be enabled, since a clock should be maintained at each tag---making synchronization service mandatory. However, the characteristics of CRFID systems expose fundamentally different challenges than battery-powered sensor networks to implement the synchronization service. The reason is mainly twofold:}

@@

The focus of this article is to answer the question of \emph{how to design a building block that synchronizes intermittently-powered CRFID systems?} To this end, we investigate the WISP platform and provide initial observations and limitations pertaining to the synchronization of these devices. In particular, the main contributions and findings of 
this article are:
%
\begin{itemize}
	\item To the best of our knowledge, this is the first study that focuses on the synchronization of tags in CRFID systems. Even though there are studies on the synchronization of multiple RFID readers, e.g.~\cite{reader_sync}, we are unaware of any study that provides synchronization of individual tags in the communication range of an RFID reader. 
	\item By addressing the aforementioned challenges, we design and implement \emph{the first synchronization primitive} for the tags, namely \emph{WISP-Sync}, inspired by the Proportional-Integral (PI) controllers~\cite{pi2015,yildirim2017adaptive}. WISP-Sync (i) 
	requires only a few computation steps to run efficiently under limited harvested energy and keeps a few variables to hold the synchronization state to recover from power interruptions with minimum overhead---\emph{addressing Challenge I}; (ii) is adaptive to react to clock instabilities in a fast manner---\emph{addressing Challenge II}.
	\item \add{Our time synchronization algorithms compensate for the clock drift and enable the first step to enable an \emph{explicit clock} for CRFID applications. Moreover, our implementations are independent of how EPC Gen2 is implemented---providing a level of portability across different platforms.}
	%
	\item We provide theoretical analysis to prove that this primitive establishes synchronization and to reveal the factors affecting its synchronization performance.
	%
	\item Evaluations in our testbed show that an average synchronization error of approximately 1\,ms can be ensured among tags with an energy overhead of 1.85\,$\mu$J per synchronization round using this  primitive.
\end{itemize}
%
We note that WISP-side implementations throughout the article are done using C and assembly language. All source codes and scripts used to generate the results in the article (including parsing, post-processing and measurement results) are available via~\url{https://goo.gl/RuWIZ2}.

@@

\add{\subsubsection{Slot and bit-level synchronization} EPC Gen2 requires a slot-level synchronization to implement slotted Aloha protocol. This is achieved by adjusting clock offsets upon reception from the reader and introducing guard times to ensure that tags communicate at the
allocated frame boundaries. It is also possible to exploit modulation properties of signals from the reader to the tags (such as bit timing in FM0 modulation) in order to get the timing measurements of the regular patterns in the signal, i.e. bit-level synchronization~\cite{liu_tcomp_2015,liu_ton_2017,liu2_ton_2017}. However, neither slot nor bit level synchronization compensate for clock drifts and output a synchronized clock value that can be accessed by the tags.}

\subsubsection{Continuously varying voltage level} In WSNs, the battery level decreases gradually that allows stable voltage levels in the short term. On the contrary, on the WISP platform, the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant drift. \add{Even though embedded platforms are equipped with voltage regulators, the output of these regulators also exhibit instability and inaccuracy with respect to the input voltage; i.e. they are not perfect and they have an input voltage transient response. Therefore, they have an output voltage variation with respect to the input voltage---still leading to clock frequency instabilities.} Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level, see Fig.~\ref{fig:frequency_dc_rf}.

@@

Even though the event-based approach is relatively lightweight and adaptive as compared to the sender-receiver synchronization, it has some limitations. First, the tag is unable to obtain an \emph{explicit reference clock value}. The only variable it tunes is the $\hat{f}_w^r$ that represents the relative frequency of the software clock with respect to the clock of the reader. Therefore, we require additional steps that will allow tags to obtain the actual reference clock value. \add{Moreover, if the tags are covered by different readers, the first step should be synchronizing the readers to the real-time, e.g. by allowing them to access time servers using internet.}
%However, to the best of our knowledge, there is no such command in the EPC Gen 2 standard that allows to send the explicit RFID reader time to the tag.
Second, WISP-Sync requires knowledge about $\tau$, which can be RFID reader dependent and requires a sniffer setup described in Section~\ref{sec:sniffer} to measure. This limitation can be addressed by employing an iterative estimation as an initial step to the WISP-Sync: the tag can start with an initial estimated $\tau$ value, and improve its estimation as it receives successive packets from the reader. After some iterations, the tag can have a rough estimate $\hat{\tau}$ that can be used on behalf of $\tau$. However, estimation errors might also perturb the synchronization accuracy. Last, the steps in Fig.~\ref{fig:event_based} present the communication scenario between one WISP and the reader. To allow other tags sniff this communication and synchronize, a broadcast primitive is crucial---in the next section we show how to overcome this limitation.

@@

%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/ksy.jpeg}}]{Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m}
%received BSc(Eng), M.Sc. and Ph.D. degrees from Ege University, İzmir, 
%Turkey in 2003, 2006 and 2012 respectively. He is  currently with Embedded Software group of TU Delft. His research interests are embedded systems, distributed algorithms,  wireless sensor networks, energy harvesting systems and computational RFIDs.
%\end{IEEEbiography}
%
%% if you will not have a photo at all:
%\begin{IEEEbiographynophoto}{Henko Aantjes}
%Biography text here.
%\end{IEEEbiographynophoto}
%
%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/przemek.jpg}}]{}
%Biography text here.	
%\end{IEEEbiography}
%
%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/amjad.jpg}}]{Amjad Yousef Majid}
%Biography text here.
%\end{IEEEbiography}
%
% insert where needed to balance the two columns on the last page with
% biographies
%\newpage

>>>>

\author{Kas{\i}m~Sinan~Y{\i}ld{\i}r{\i}m, 
	Henko~Aantjes, 
	Przemys{\l}a~Pawe{\l}czak~\IEEEmembership{Member,~IEEE,} 
	and~Amjad~Yousef~Majid
	%
	\IEEEcompsocitemizethanks{
		\IEEEcompsocthanksitem A preliminary version of this article was presented at the HLPC 2016 Workshop, Atlanta, GA, USA, April 2, 2016~\cite{hlpc2016}.
		\IEEEcompsocthanksitem The authors are with the Embedded Software Group, Delft University of Technology, Mekelweg 4, 2628 CD Delft, The Netherlands (e-mail: \{k.s.yildirim, p.pawelczak, a.y.majid\}@tudelft.nl, h.aantjes@student.tudelft.nl). 
		\IEEEcompsocthanksitem Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m is also with the Computer Engineering Department of Ege University, Izmir, Turkey.
	}
%	\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}
}

@@

As of now, battery-less networks are in the form of one-hop architecture in which the CRFID tags communicate only with the RFID reader~\cite{wisp,booksmith2013,wisent:2016}. Therefore, CRFID tags cannot operate in a multi-hop manner. However, recent advancements in the wireless communication enabled tag-to-tag communication, e.g.~\cite{liu2013ambient}.  We anticipate that this communication capability will lead up multi-hop CRFID networks that will demand their own implementation of basic sensor network building blocks~\cite{levis:tinyos} with the aim of replacing the existing battery-powered sensing and data collection applications with their battery-less counterparts.

Since we are not aware of any network infrastructure demonstrating neither multi-hop communication nor multi-hop synchronization, our focus in this article are the CRFID networks where several readers cover a tag population simultaneously in a single hop. A representative application is the greenhouse monitoring where several tags are powered by multiple readers, see Figure~\ref{fig:motivation}. In this application, the tags sense the environment and timestamp their temperature readings. When the readers request their temperature readings, the tags transmit their timestamps together with the collected temperature data. In order to make inferences on the collected temperature readings, timestamps should be consistent so that the readers can provide a temporal ordering of the readings. Therefore, tags require a common time notion to timestamp their readers---creating a demand for explicit time synchronization. In order to achieve synchronization, readers should first acquire the global time, e.g. by accessing time servers using the Internet; then individual tags should be synchronized to the individual readers.

@@

\begin{figure}
	\centering
	\includegraphics[width=0.75\columnwidth]{figures/scenario.pdf}
	\caption{\label{fig:motivation}A representative greenhouse monitoring application using CRFIDs. Two readers are deployed to provide power to the individual tags and to collect their periodic sensor readings.}
\end{figure}

@@

The EPC Gen2 protocol uses slotted Aloha that performs a form of synchronization between the reader and the tags: by adjusting clock offset differences and introducing guard times, tags are ensured to communicate at the allocated slot boundaries. However, this method does not compensate for clock drifts and output a synchronized clock value that can be accessed by the tags to timestamp (and make inference about) the collected information, e.g. from the sensors attached to tags. In consequence, applications requiring time notion cannot be enabled, since a clock should be maintained at each tag---making synchronization service mandatory. However, the characteristics of CRFID systems expose fundamentally different challenges than battery-powered sensor networks to implement the synchronization service. The reason is mainly twofold:

@@

The focus of this article is to answer the question of \emph{how to design a building block that synchronizes intermittently-powered CRFID systems?} To this end, we investigate the WISP platform and provide initial observations and limitations pertaining to the synchronization of these devices. In particular, the main contributions and findings of 
this article are:
%
\begin{itemize}
	\item To the best of our knowledge, this is the first study that focuses on the synchronization of tags in CRFID systems. Even though there are studies on the synchronization of multiple RFID readers, e.g.~\cite{reader_sync}, we are unaware of any study that provides synchronization of individual tags in the communication range of an RFID reader. 
	\item By addressing the aforementioned challenges, we design and implement \emph{the first synchronization primitive} for the tags, namely \emph{WISP-Sync}, inspired by the Proportional-Integral (PI) controllers~\cite{pi2015,yildirim2017adaptive}. WISP-Sync (i) 
	requires only a few computation steps to run efficiently under limited harvested energy and keeps a few variables to hold the synchronization state to recover from power interruptions with minimum overhead---\emph{addressing Challenge I}; (ii) is adaptive to react to clock instabilities in a fast manner---\emph{addressing Challenge II}.
	\item Our time synchronization algorithms compensate for the clock drift and enable the first step to enable an \emph{explicit clock} for CRFID applications. Moreover, our implementations are independent of how EPC Gen2 is implemented---providing a level of portability across different platforms.
	%
	\item We provide theoretical analysis to prove that this primitive establishes synchronization and to reveal the factors affecting its synchronization performance.
	%
	\item Evaluations in our testbed show that an average synchronization error of approximately 1\,ms can be ensured among tags with an energy overhead of 1.85\,$\mu$J per synchronization round using this  primitive.
\end{itemize}
%
We note that WISP-side implementations throughout the article are done using C and assembly language. All source codes and scripts used to generate the results in the article (including parsing, post-processing and measurement results) are available via~\url{https://goo.gl/RuWIZ2}.

@@

\subsubsection{Slot and bit-level synchronization} EPC Gen2 requires a slot-level synchronization to implement slotted Aloha protocol. This is achieved by adjusting clock offsets upon reception from the reader and introducing guard times to ensure that tags communicate at the
allocated frame boundaries. It is also possible to exploit modulation properties of signals from the reader to the tags (such as bit timing in FM0 modulation) in order to get the timing measurements of the regular patterns in the signal, i.e. bit-level synchronization~\cite{liu_tcomp_2015,liu_ton_2017,liu2_ton_2017}. However, neither slot nor bit level synchronization compensate for clock drifts and output a synchronized clock value that can be accessed by the tags.

\subsubsection{Continuously varying voltage level} In WSNs, the battery level decreases gradually that allows stable voltage levels in the short term. On the contrary, on the WISP platform, the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant drift. Even though embedded platforms are equipped with voltage regulators, the output of these regulators also exhibit instability and inaccuracy with respect to the input voltage; i.e. they are not perfect and they have an input voltage transient response. Therefore, they have an output voltage variation with respect to the input voltage---still leading to clock frequency instabilities. Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level, see Fig.~\ref{fig:frequency_dc_rf}.

@@

Even though the event-based approach is relatively lightweight and adaptive as compared to the sender-receiver synchronization, it has some limitations. First, the tag is unable to obtain an \emph{explicit reference clock value}. The only variable it tunes is the $\hat{f}_w^r$ that represents the relative frequency of the software clock with respect to the clock of the reader. Therefore, we require additional steps that will allow tags to obtain the actual reference clock value. Moreover, if the tags are covered by different readers, the first step should be synchronizing the readers to the real-time, e.g. by allowing them to access time servers using internet.
%However, to the best of our knowledge, there is no such command in the EPC Gen 2 standard that allows to send the explicit RFID reader time to the tag.
Second, WISP-Sync requires knowledge about $\tau$, which can be RFID reader dependent and requires a sniffer setup described in Section~\ref{sec:sniffer} to measure. This limitation can be addressed by employing an iterative estimation as an initial step to the WISP-Sync: the tag can start with an initial estimated $\tau$ value, and improve its estimation as it receives successive packets from the reader. After some iterations, the tag can have a rough estimate $\hat{\tau}$ that can be used on behalf of $\tau$. However, estimation errors might also perturb the synchronization accuracy. Last, the steps in Fig.~\ref{fig:event_based} present the communication scenario between one WISP and the reader. To allow other tags sniff this communication and synchronize, a broadcast primitive is crucial---in the next section we show how to overcome this limitation.

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/ksy.jpeg}}]{Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m}
received his Ph.D. degree in Computer Engineering from Ege University 
in 2012. He was a visiting scholar at the Department of Information Engineering, University of Padova in 2013. Between 2015-2017, he was a post-doctoral researcher at Embedded Software Group, Delft University of Technology. Currently, he is an assistant professor at the Department of Computer Engineering, Ege University and also a research affiliate with the Embedded Software Group at Delft University of Technology. His research interests are embedded systems and software, distributed algorithms,  wireless sensor networks, energy harvesting systems and computational RFIDs.
\end{IEEEbiography}

% if you will not have a photo at all:
\begin{IEEEbiographynophoto}{Henko Aantjes}
Biography text here.
\end{IEEEbiographynophoto}
%
\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/przemek.jpg}}]{}
Biography text here.	
\end{IEEEbiography}
%
\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/amjad.jpg}}]{Amjad Yousef Majid}
Biography text here.
\end{IEEEbiography}
%
% insert where needed to balance the two columns on the last page with
% biographies
%\newpage
